Common sense approaches to sharing tabular data alongside publication

### 2021-09-17 new-paper.Rmd
address R2-5
added bib and fixed R1-1 and R1-3

While there has been a growing trend of sharing data alongside publications, the status quo of data sharing in science is not good. Rowhani-Farid et al. [@Rowhani-Farid2016] showed that 7 of 157 (4.5%) sampled papers in the *British Medical Journal* shared data, and only 14% of articles claiming open sharing of data actually shared their data. Similarly, Stodden et al. [@Stodden2018] showed that only 36% of papers published in the journal *Science* shared data, despite the journal's data sharing policy. Similar findings have been made for PLOS and BMC journals [@Colavizza2020]. There are numerous examples of similar problems in other fields, such as psychology [@vanpaemel2015] and biology [@schofield2009]. The reality is that while data sharing is increasing in many fields, such as ecology and genomics, the amount of data that can be reused is likely lower. To make shared data useful, we need to realign incentives for researchers.

@@

Journals provide data papers in a formal publication explaining data that is archived. This is a familiar avenue for researchers to receive credit and citation. Writing a data paper can be very useful for researchers, but, data papers lack guidelines on how to structure data for effective reuse and sharing. Partner with Dryad to link archived data with research papers. Dryad will also provide professionals to curate your data, thereby improving usability. While writing a data paper can be very useful for researchers, journals that publish data papers lack guidelines on how to structure data for effective reuse and sharing.

@@

One critical advantage of depositing data in a trusted repository (Section \@ref(avenues)) is that it comes with many benefits, such as preservation, indexing,  discoverability, and most importantly, citation with a persistent identifier. Authors can take an extra step by including their own CITATION file, in which the data and associated methods are described in detail. When citing data, you should cite the preferred CITATION file in the first instance, else, if there is a data publication, (for example in _Scientific Data_), cite this, and if none of those apply, cite the repository in which the data is deposited (e.g., a Zenodo repository).

>>>>

While there has been a growing trend of sharing data alongside publications, the status quo of data sharing in science is not good. In 2016, Rowhani-Farid et al. [@Rowhani-Farid2016] showed that 7 of 157 (4.5%) sampled papers in the *British Medical Journal* shared data, and only 14% of articles claiming open sharing of data actually shared their data. Similarly, in 2018, Stodden et al. [@Stodden2018] showed that only 36% of papers published in the journal *Science* shared data, despite the journal's data sharing policy. Similar findings have been made for PLOS and BMC journals [@Colavizza2020]. There are numerous examples of similar problems in other fields, such as psychology [@vanpaemel2015] and biology [@schofield2009]. The reality is that while data sharing is increasing in many fields, such as ecology and genomics, the amount of data that can be reused is likely lower. To make shared data useful, we need to realign incentives for researchers.

@@

Journals provide data papers in a formal publication explaining data that is archived. This is a familiar avenue for researchers to receive credit and citation. Writing a data paper can be very useful for researchers, but, data papers lack guidelines on how to structure data for effective reuse and sharing. Partner with Dryad to link archived data with research papers. We recommend partnering with best practice data repositories that emphasize curation, like Dryad. This way the data will have a long-term archiving and will also have professional curation to ensure that the published data are usable and have enough metadata to be understood. While writing a data paper can be very useful for researchers, journals that publish data papers lack guidelines on how to structure data for effective reuse and sharing. Data journals should routinely be partnering with best practice repositories to best support the underlying data.

@@

One critical advantage of depositing data in a trusted repository (Section \@ref(avenues)) is that it comes with many benefits, such as preservation, indexing,  discoverability, and most importantly, citation with a persistent identifier. Authors should add a CITATION file when possible but the real burden lies with publishers to ensure that these citations are marked up appropriately to allow for the building of open data metrics [@lowenberg_2021, @lowenberg_daniella_2021b].

When citing data, you should cite the preferred CITATION file in the first instance, else, a data publication, (for example in _Scientific Data, Ecology_), and if none of those apply, cite the repository in which the data is deposited (e.g., a Zenodo repository).
